---
title: "Swiss Fertility: A Correlation Analysis"
subtitle: "Associations among socio-economic indicators in Swiss districts"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
    code-fold: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Packages (install only if missing)
needed <- c("readr", "dplyr", "tidyr", "ggplot2", "knitr")
missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

## Introduction

The classic Swiss fertility data describe 47 French-speaking districts of
Switzerland around 1888. Each district is characterised by six quantitative
measures: fertility, the proportion of males in agriculture, the results of a
military examination, education, the percentage of Catholics, and infant
mortality.

This document loads the data directly from a remote CSV, computes the Pearson
correlation matrix among the numeric variables, and displays it as an
annotated heatmap.

## Load the data

The dataset is read straight from GitHub — no local copy is required.

```{r}
#| label: load-data

data_url <- "https://raw.githubusercontent.com/Marshall-Soc/css_bootcamp/main/data/swiss.csv"

swiss <- read_csv(data_url, show_col_types = FALSE)

# The first column holds district names but arrives with an empty header.
names(swiss)[1] <- "District"

glimpse(swiss)
```

```{r}
#| label: preview
kable(head(swiss, 8), digits = 2, caption = "First six districts of the Swiss data.")
```

## Correlation matrix

We correlate every numeric column with every other. The leading `District`
column is dropped first so that only quantitative variables enter the matrix.

```{r}
#| label: corr-matrix

swiss_num <- swiss %>%
  select(where(is.numeric))

corr_mat <- cor(swiss_num, use = "complete.obs")
round(corr_mat, 2) %>%
  kable(caption = "Pearson correlation matrix (rounded to 2 decimals).")
```

### Strongest associations

```{r}
#| label: extremes

# Rank off-diagonal pairs by absolute correlation
pairs <- as.data.frame(as.table(corr_mat)) %>%
  filter(Var1 != Var2) %>%
  mutate(abs_r = abs(Freq)) %>%
  arrange(desc(abs_r)) %>%
  slice_head(n = 5)

kable(
  pairs,
  col.names = c("Variable 1", "Variable 2", "r", "|r|"),
  digits = 3,
  row.names = FALSE,
  caption = "Five strongest off-diagonal correlations."
)
```

## Annotated correlation heatmap

Each cell is coloured by the correlation coefficient and labelled with its
value, so the plot can be read without consulting a separate legend of numbers.

```{r}
#| label: heatmap
#| fig-width: 8
#| fig-height: 7
#| fig-align: center

# Preserve the matrix's variable ordering on both axes
var_levels <- colnames(corr_mat)

corr_df <- as.data.frame(corr_mat) %>%
  mutate(Variable1 = factor(rownames(corr_mat), levels = var_levels)) %>%
  pivot_longer(
    cols = -Variable1,
    names_to = "Variable2",
    values_to = "r"
  ) %>%
  mutate(Variable2 = factor(Variable2, levels = var_levels))

ggplot(corr_df, aes(x = Variable2, y = Variable1, fill = r)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 4.2, family = "mono") +
  scale_fill_gradient2(
    low    = "#2166AC",
    mid    = "white",
    high   = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1),
    name   = "Pearson r"
  ) +
  coord_fixed() +
  labs(
    title    = "Correlation heatmap of Swiss district indicators",
    subtitle = "Blue = negative association, red = positive association",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y  = element_text(face = "bold"),
    panel.grid   = element_blank(),
    plot.title   = element_text(face = "bold")
  )
```

## Interpretation

* **Fertility and education** are strongly negatively correlated: districts
  with more schooling tended to have lower fertility.
* **Fertility and Catholic** show a clear positive association, while
  **agriculture** is negatively related to education and examination results.
* **Infant mortality** is largely disconnected from the socio-economic
  indicators: its correlations with agriculture, examination, and education are
  all within $\pm 0.11$ of zero. Its one notable association is a moderate
  positive link with fertility ($r = 0.42$).

Correlation does not imply causation, and these are district-level (ecological)
quantities, so the coefficients should be read as descriptive summaries of
association rather than as causal effects.

## Session information

```{r}
#| label: session
sessionInfo()
```
